Supplementary material for Sequential rank agreement methods for comparison of ranked lists

### 2018-01-24 latex/biosts3-supp.tex
Updatedd version

\documentclass[12pt,a4paper]{article}
\pdfoutput=1
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
%\usepackage[noae]{Sweave}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}

@@

\newcommand{\mcomment}[1]{\textcolor{red}{\textbf{#1}}}

@@

\usepackage[amsthm,thmmarks]{ntheorem}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\newtheorem{example}{Example}
\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

@@

\title{Sequential rank agreement methods for comparison of ranked lists}
\author{Claus Thorn Ekstrøm}
\author{Thomas Alexander Gerds}
\author{Andreas Kryger Jensen}
%\author{Kasper Brink-Jensen}
\affil{Biostatistics, University of Copenhagen}
\renewcommand{\topfraction}{.85}
\renewcommand{\bottomfraction}{.7}
\renewcommand{\textfraction}{.15}
\renewcommand{\floatpagefraction}{.66}
\renewcommand{\dbltopfraction}{.66}
\renewcommand{\dblfloatpagefraction}{.66}
\setcounter{topnumber}{9}
\setcounter{bottomnumber}{9}
\setcounter{totalnumber}{20}
\setcounter{dbltopnumber}{9}

@@

@@

@@

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete. The approach
  can identify changepoints in the agreement of the lists, and the
  sequential changes of agreement as a function of the depth of the
  lists can be compared graphically to a permutation based reference
  set or evaluated using analytical asymptotic results.  The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies where we assess
  the within and between agreement of different statistical
  classification methods.
\end{abstract}

Key words: sequential rank agreement, partial ranking, order
statistic, gene rankings, methods comparison, variable selection,
permutation

\section{Introduction}

Ranking of items or results is common in scientific research and
ranked lists occur naturally as the result of many statistical
applications. Regression methods rank predictor variables according to
magnitude of their association with an outcome, prediction models rank
subjects according to their risk of an event, and genetic studies rank
genes according to their difference in gene expression levels across
samples.

When several rankings of the same items are available, a common
research question is to what extent they agree. In particular, is it
possible to identify an optimal rank until which the lists agree on
the items?  A typical situation arises in high-dimensional genomics
studies when several analysis methods are applied to rank a list of
genes according to their association with a phenotype, treatment
effect or other outcome. A measure of agreement of gene rankings
obtained by different methods could often help to identify which genes
are worth to pursue in further experiments.

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists and they both provide a
single measure of the overall distance. \citet{Shieh1998} proposed a
weighted version of Kendall's $\tau$ where each pair of rankings can
be assigned different weights, and \citet{Yilmaz2008} proposed the
$\tau_{ap}$ which places higher emphasis on the top of the lists.
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the similarity measure.

Other recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list and therefore \citet{Fagin2003} and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists. 
% One proposal is denoted the average overlap (AO) and used for
% comparison later on. 
Specifically, \citet{Webber2010} define their rank-biased overlap
(RBO) by weighting with a converging series to ensure that the top is
weighted higher than the potentially non-informative bottom of the
lists.  It is possible to use the existing methods to calculate
agreement of lists until a given depth, i.e., limited to the $d$ items
of each list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

Very recently, \citet{hall:schi:2012} proposed a method for comparing
pairwise rankings and derived the asymptotic distribution of the
endpoint where the two ranked lists no longer are in agreement. Their
approach was based on anchoring one of the two lists and subsequently
generating a sequence of 0s and 1s depending on whether the ranks in
the second list was close to the rank from the anchored
list. \citet{samp:verd:2013} followed up on this idea for pairwise
comparison of lists but used penalties based on a truncated geometric
probability instead of a 0-1 process and they evaluated the
distribution of the endpoint of agreement by computational approaches.
The asymptotic distribution in the \citet{hall:schi:2012} paper is
based on letting the number of \emph{lists} increase to infinity which
is a situation that is only relevant in special cases, whereas the
simulation-based null distribution approach of \citet{samp:verd:2013}
is guaranteed to work also for small samples.

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could essentially be used. Our proposed approach allows us to
compare multiple lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, it places
higher weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
changepoint, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide asymptotical and randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete lists. In section 3 we present and discuss
approaches to evaluate the results obtained from sequential rank
agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies and compare our method to the method of
\citet{hall:schi:2012} in a small sample simulation study before we
discuss the findings along with possible extensions. The approaches
presented in this manuscript are available in the R package
\texttt{SuperRanker} which can be found on CRAN.
% chunk 1

\section{Methods}

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to
the set of items $X$.  Panels (a) and (b) of Table~\ref{tab:example}
show a schematic example of these mappings. Thus if
$R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked first in list $l$
and similarly $R_l(X_{34})=1$.

In all what follows we consider a fixed set of items and consider the
ranking function to be a random variable. Thus, let
$R_1(X),\dots,R_L(X)$, $L\geq2$, be a sample of $L$ independent
identically distributed draws from an unknown probability distribution
function $Q$. One aim is then to test how much $Q$ resembles the
uniform distribution which assigns probability $1/P!$ to each of the
$P!$ different possible rankings.

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked lists of
    items for each of three lists, (b) presents the ranks obtained by
    each item in each of the three lists and (c) shows the cumulative
    set of items up to a given depth in the three lists.}
\begin{center}
\footnotesize{
  \begin{subtable}{4.5cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $R^{-1}_1$ & $R^{-1}_2$ & $R^{-1}_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%%
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
}
\end{center}
\label{tab:example}
\end{table}

The agreement among the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_p)-\E_Q R(X_p)\right)^2\right]\\
   &= \sum_{r\in\Pi} \left(r(X_p)-\E_Q R(X_p)\right)^2 Q(r),\nn
\end{align}
where $\Pi$ is the set of all permutations of $X$, $Q$ is a
probability mass function on $\Pi$, and
$\E_Q R(X_p)=\sum_{r\in\Pi} r(X_p)Q(r)$. The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{align}
S(d) &= \left\{X_p; \left(\sum_{r\in\Pi} 1\left(r(X_p)\le d\right) Q(r)\right) >0\right\}\label{eq:sumset}
\end{align}
where $1(\cdot)$ denotes the indicator function.
The empirical counterpart is the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{L}(d) = \left\{X_p; \left(\frac{1}{L}\sum_{l=1}^L 1\left(R_l(X_p)\le d\right)\right)>0\right\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.

We define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S(d)$:
\begin{align}
  \textrm{sra}(d) = \frac{1}{|S(d)|}\sum_{p \in S(d)}A(X_p)\label{def:sra}
\end{align}
where $|S(d)|$ is the cardinality of the set $S(d)$.
The empirical counterpart is equivalently given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{L}(d)\}}(L-1)\widehat{A}_L(X_p)}{(L-1)|\widehat {S}_{L}(d)|}\label{def:sraEst}
\end{equation}

Values of \textrm{sra} close to zero suggests that the lists agree on
the rankings while larger values suggest disagreement.  Just like in
method comparison studies we prefer smaller limits of agreement
because that means that the differences between the rankings is
small. If the ranked lists are identical then the value of sequential
rank agreement will be zero for all position or depths $d$ down the
lists.  The sequential rank agreement is equivalent to the pooled
variance of the items found in $S(d)$, and it can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.
For an easier interpretation we suggest to plot
$\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding to the pooled
standard deviation.

\subsection{Agreement among fully observed lists}
\label{sec:amfol}

The simplest case occurs when all $L$ lists are fully observed, \ie, when
we observe the rank of all $P$ items for all $L$ lists. Fully
observed ranked lists are common and arise, for example, when
different statistical analysis methods are applied to a single dataset
to produce lists of predictors ranked according to their importance,
or if the same analysis method is applied to data from different
populations.

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in the top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal unequal variances two-sample $t$ tests,
marginal logistic regression analyses, elastic net logistic regression
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to minimum $p$ value, for
logistic regression the genes were ordered by size of the
corresponding coefficients (after standardization), and MIC was
ordered by absolute correlation which resulted in the top rankings
seen in Table~\ref{tab1}.

The sequential rank agreement curve shown in the two left plots of
Figure~\ref{fig:example1} roughly indicates the average distance in
ranks for the genes found among the first $d$ positions in the $4$
lists.  It can be seen that the sequential rank agreement limits are
better towards the top of the lists (smaller values on the $y$ axis
corresponds to \emph{better} agreement) than towards the bottom of the
lists. It is clear from the curve in Figure~\ref{fig:example1} that
there is a substantial drop in agreement (higher sra) even after the
first depth. Thus, if we were to restrict attention to a small set of
predictors then we would focus on predictors 2124, 829, and 378 as
seen in Table~\ref{tab1}.

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{List of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method. Only the top 10 ranks are shown in the table.} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & Welsh's $t$ & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

% chunks 2,3,4,5,6,7

\begin{figure}[htbp]
   \begin{center}
 \includegraphics[width=.49\textwidth]{paper-fig1a}
 \includegraphics[width=.49\textwidth]{paper-fig2a}
 \includegraphics[width=.49\textwidth]{paper-fig1b}
 \includegraphics[width=.49\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data (black
   line). Right panels: Corresponding sequential rank agreement for
   the same data but where only the top 20 ranked items are available
   and the rank of the remaining items are not available. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

% If the lists agree on all items then the sequential rank agreement is
% zero for all depths. 
 Generally, if there are changes in the level of
 rank agreement then this suggests that there are sets of items that
 are ranked similarly in all lists and other sets of items that have
 been assigned vastly different ranks in the lists. For example, in gene
 association studies we generally expect the agreement among the lists
 to be better towards the top of the lists and worse towards the end
 of the lists. In this case the sequential rank agreement curve will
 start at a low level and then increase until it levels off as seen in
 Figure~\ref{fig:example1}.

\subsection{Analysis of incomplete lists}
Incomplete or partial lists are a common occurrence that arise, for
example, in case of missing data (items), when comparing top $d$ list
results from publications, or when some methods only rank a subset of
the items. For example, penalized regression based on the Lasso
provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering.  Incomplete lists also occur if the analyst
restricts attention to the ranks of items that have been found to be
statistically significant.

Sequential rank agreement can be generalized to incomplete lists in
the following way. Let $\Lambda_l\subset X$ be the subset of $d_l$
items that have been ranked highest in list $l$.  The case where all
lists are incomplete at the same depth $d$ corresponds to
$d_1=\cdots=d_L=d$. For incomplete lists the rank function becomes
\begin{equation}
	\widetilde R_l(X_p) = \begin{cases}
	     \{R_l(X_p)\} & \text{for } X_p\in \Lambda_l\\
	     \{d_l+1,\dots,P\} & \text{for} X_p \not\in \Lambda_l
        \end{cases}
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of incomplete lists because the exact rank
for some items will be unknown. Also, recall that the rankings within
a single list are not independent since each rank must appear exactly
once in each list. Thus, we cannot simply assign the same number
(\eg, the mean of the unassigned ranks) to the missing items since
that would result in \emph{less} variation of the ranks and hence less
variation of the agreement, and it would artificially introduce a
(downward) bias of agreement for items that are missing in multiple
lists.

Instead we randomize the ranks $\{d_{l}+1,\dots,P\}$ to the items that
do not occur in list $\Lambda_l$. One realization of the $L$ rankings
of the set $X$ is obtained by randomizing the missing items of each
list. By randomizing a large number of times we can compute
\eqref{def:sra} for each realization, and then compute the sequential
rank agreement as the pointwise (for each depth) average of the rank
agreements. The algorithm is described in detail in
Algorithm~\ref{sra-algorithm}.

\begin{algorithm}
\caption{Sequential rank agreement algorithm for incomplete lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
  \Procedure{Incomplete list rank agreement}{} \State Let $B$ be the number
  of permutations to use \For{each $b \in B$} \For{each list
    $l \in L$}
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute
    the unassigned ranks, $\{d_l+1, \ldots, P\}$, and assign them
    randomly to the items \emph{not} found in the list, i.e.,
    $\Lambda^\complement=X\setminus\Lambda_l$. Combine the
    result with $\Lambda_l$ to fill out the list.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

The proposed approach is based on two assumptions: 1) that the most
interesting items are found in the top of the lists, and 2) that the
ranks that are missing from the lists provide so little information that it is reasonable
to assume that they can be represented by a random order. The first
assumption is justifiable because we have already accepted that it is
reasonable to rank the items in the first place. The second assumption
is fair in the light of the first assumption provided that we have a
``sufficiently large'' part of the top of the lists available.

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the number of items without ranks available is
low. Generally, without additional prior knowledge about the
underlying distribution of the ranks it seems reasonable to restrict
attention of the sequential rank agreement to lists no longer than the
longest of the $L$ lists examined.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For incomplete ranked lists we also
expect a changepoint around the depth where the lists are
become incomplete. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

The two right-most plots of Figure~\ref{fig:example1} show the impact
of restricting the Golub data such that only top-20 lists are
available instead of full lists of length 1000 (20 was chosen to
resemble the list lengths that might be published in a
manuscript). The sequential rank agreement increases much quicker
because the incomplete lists introduce more noise in the estimation of
the agreement, but it is still possible to see that the top of the
list has a sequantial rank agreement that is not substantially
different from the full lists.

% \subsection{Comparison to average overlap}
% The average overlap is a measure that is widely used in comparisons of
% two fully observed ranked lists \citep{Fagin2003,Webber2010} and the
% idea behind the average overlap closely resembles the sequential rank
% agreement. In this section we present an extension of the average
% overlap to more that two lists and compare this to sequential rank
% agreement.

% The overlap among $L$ lists observed until depth $d$ is defined as the
% proportion of items that are found in all $L$ lists compared to the
% possible number of items found in all lists observed until depth $d$:
% \begin{align}
%   O(d) = \frac{| \cap_{l=1}^L \{R_l^{-1}(r); r\leq d \} |}{d}\label{def:overlap}
% \end{align}
% Typically the overlap is only considered for pairwise comparisons
% \citep{Bar-Ilan2006,Boulesteix2009} but formula \ref{def:overlap}
% is easily extended to the situation where $L>2$ as shown above.

% The average overlap at depth $d$ is defined as the average of the
% overlaps until depth $d$,
% \begin{equation}
% AO(d) = \frac1d\sum_{i=1}^d O(d).
% \end{equation}
% This definition directly ensures that further emphasis is put on items
% in the top of the lists since their item specific overlap contributes
% to the calculation of average overlap at higher depths.

% There are three main drawbacks of the average overlap method: First it
% is highly sensitive to single items in the top of one the lists.  The
% average overlap profile may change dramatically depending on the
% presence or absence of similar items in the beginning of the
% lists. For example, if all lists have the same item at the top then
% the average overlap attains its maximum at 1, but if just one of the
% lists differ then the overage overlap starts at its minimum of
% 0. Secondly, when the number of lists, $L$, is high then it might be
% difficult to obtain a non-zero overlap (and hence a non-zero average
% overlap) at the beginning of the lists because the overlap will be
% zero until an item is present in all $L$ lists. Finally, the average
% overlap has an interpretation in terms of moving averages of
% percentages (of $L$ sets) which is somewhat less intuitive than our
% proposed sequential rank distance.

% In the following we compare the average overlap and the sequential
% rank agreement using the analysis of the Golub data
% \citep{Golub1999}. We consider a typical application of the average
% overlap and thus restrict attention to only two gene rankings obtained
% by the $t$ tests and logistic regression. These two marginal modeling
% approaches should result in very similar gene rankings. The first two
% columns of Table~\ref{tab1} show the actual top-10 ranking of the
% predictors.

% % chunks 8,9

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=.49\textwidth]{paper-fig-ao1}
% \includegraphics[width=.49\textwidth]{paper-fig-ao2}
% \end{center}
%  \caption{Average overlap (left figure) and sequential rank agreement
%  (right figure) for comparing two different analysis methods
%  (marginal $t$ test and logistic regression)
%     applied to the Golub data. The black lines are based on the full
%     set of predictors while the red lines have the two highest
%     associated predictors removed
%     from the data before computing the average overlap and sequential
%     rank agreement.}
%   \label{fig:case1}
%  \end{figure}

%  The black lines in Figure~\ref{fig:case1} show the average overlap
%  and sequential rank agreement for these data. It is clear from both
%  plots that there is perfect agreement towards the top of the ranked
%  lists: the average overlap is 1 and the sequential rank agreement is
%  zero.  However, if we remove those two items from the data (i.e.,
%  gene/predictor 2124 and 896) and redo the analyses then we get
%  substantially different curves for the average overlap but not for
%  sequential rank agreement (red lines in Figure~\ref{fig:case1}). For
%  the sequential rank agreement we get roughly the same estimate of
%  agreement from rank 3 as we did from the full dataset. This indicates
%  that sequential rank is more robust against small perturbations of
%  the data. For the average overlap the curves completely change when
%  these two items are removed and which would lead to a different
%  conclusion about the agreement among the items from rank 3 downwards.

\section{Evaluating sequential rank agreement}
To interpret the sequential rank agreement values we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

The first reference hypothesis is
\begin{eqnarray}
H_0  & : &  \text{The list rankings correspond to completely randomly}\label{eq:permutationHypothesis}\\
       &  & \text{permuted lists}\nn
\end{eqnarray}
which not only assumes that there is no information in the data on
which the rankings are based but also that the methods used to provide
the rankings are completely independent.

Alternatively, we can remove the restriction on the independence among
the methods used to generate the individuals rankings in the $L$ lists
by only requiring that there is no information contained in the ranking but that
the rankings are all based on applying the method/approaches to the
same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

This alternative null hypothesis addresses the fact that some ranking
methods are more likely to provide similar rankings of the same data
because the ranking methods focus on the same features of the data
rather than because of any information contained in the data.

\subsection{Permutation-based inference}

$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the fully observed case each experiment
contains $L$ lists of random permutations of the items in $X$. For the
incomplete case we first permute the items $X_1,\dots,X_P$ and then assign missing ranks
for list $l$ from  $d_l$ to $P$ (\ie, each list so it has the same number of
observed rankings as was observed for list $l$ in the original
dataset). The sequential rank agreement curve from the original
lists can then be compared to, say, the pointwise 95\% quantiles of
the observed rank agreements obtained under $H_0$.

To obtain the distribution under $\widetilde H_0$ the idea is to
repeat the ranking procedures for unassociated data many times.  Thus,
we first permute the outcome variable of the dataset on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
dataset to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. This randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\widetilde H_0$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $\widetilde H_0$ then this suggests that
the original ranked lists agree \emph{more} than what we would expect
in data with no information, and therefore that the information in the
lists is significantly more in agreement than what would be expected.

Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\widetilde H_0$ each based
on $400$ permutations of the Golub data described in Section
\ref{sec:amfol}. Not surprisingly, the sequential rank agreement under
$\widetilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank) similar predictors even if there are only spurious
associations. The two bottom panels in Figure~\ref{fig:example1} also
indicate that the observed sequential rank agreement (the black line)
is better than what would be expected by chance for data that contain
no information since it lies below the reference areas. The incomplete
data also suggests that there may be at most 1 or 2 ranked items
towards the top of the lists that yield a result better than what
would be expected (the bottom-right plot).

It is important to stress that neither $H_0$ nor $\widetilde H_0$ are
related to questions regarding the association between the outcome and
the predictors in the dataset. Both hypotheses are purely considering
how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

\subsection{Asymptotic inference of change in agreement}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depth which
satisfies a changepoint criterion since that corresponds to a change
in agreement among the list ranks. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As in \cite{hall:schi:2012} we consider an infinite set
of lists and study the asymptotic behaviour for $L\to\infty$.
However, we do not allow the length of the lists to change with $L$
because in most applications the lengths of the lists are fixed.

We start by showing that $\widehat{\textrm{sra}}_L$ is a consistent
estimator of $\textrm{sra}$ for $L \rightarrow \infty$. The result is
stated in Theorem \ref{thm:consistency}.

% below and requires the
% following assumptions.
% \begin{assumption}
% We assume that:
% \begin{enumerate}[label=\Alph*)]
% \item{The $L$ lists are independent and identically distributed
    % samples from a probability distribution $Q$ on the set of lists.}
% \item{$\E_Q[R(X_p)^2] < \infty$ uniformly in $p$.}
% \item{$|S(d)| > 0$ uniformly in $d$}.
% \end{enumerate}
% \end{assumption}
% We note that assumption $C$ implies that for every list depth $d$
% there must be a non-zero probability of ... under the probability
% measure $Q$.

\begin{theorem}
  Assume that $\{R_l(X)\}_{l=1}^L$ are independent draws from a
  probability distribution $Q$ on the set of lists $\Pi$. Then,
  $\left\|\widehat{\textrm{sra}}_L - \textrm{sra}\right\|_\infty =
  o_P(1)$.
\label{thm:consistency}
\begin{proof}
See Appendix \ref{sec:appA}.
\end{proof}
\end{theorem}

>>>>

\documentclass[oupdraft]{bio}
%\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
% \documentclass[12pt,a4paper]{article}
\pdfoutput=1
\usepackage[utf8]{inputenc}
%\usepackage[]{graphicx}
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage[]{color}
%\usepackage{subcaption}
% \usepackage{natbib}
%\usepackage{xspace}
%\usepackage{calc}
% \usepackage[noae]{Sweave}
%\usepackage{algorithm}
%\usepackage{algpseudocode}
% \usepackage{enumitem}

% \usepackage{authblk}

% \usepackage[nolists,nomarkers]{endfloat}

%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
%\newcommand{\ie}{\emph{i.e.}\@\xspace}
%\newcommand{\eg}{\emph{e.g.}\@\xspace}

@@

% \newcommand{\mcomment}[1]{\textcolor{red}{\textbf{#1}}}

@@

%\usepackage[amsthm,thmmarks]{ntheorem}
% \theoremstyle{plain}
% \newtheorem{theorem}{Theorem}
%\newtheorem{definition}{Definition}
%\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}[section]
%\newtheorem{lemma}{Lemma}
%\newtheorem{assumption}{Assumption}
%\newtheorem{example}{Example}
%\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

@@

%\title{Sequential rank agreement methods for comparison of ranked lists}
%\author{Claus Thorn Ekstrøm}
%\author{Thomas Alexander Gerds}
%\author{Andreas Kryger Jensen}
%\author{Kasper Brink-Jensen}
%\affil{Biostatistics, University of Copenhagen}
%\renewcommand{\topfraction}{.85}
%\renewcommand{\bottomfraction}{.7}
%\renewcommand{\textfraction}{.15}
%\renewcommand{\floatpagefraction}{.66}
%\renewcommand{\dbltopfraction}{.66}
%\renewcommand{\dblfloatpagefraction}{.66}
%\setcounter{topnumber}{9}
%\setcounter{bottomnumber}{9}
%\setcounter{totalnumber}{20}
%\setcounter{dbltopnumber}{9}

\renewcommand{\topfraction}{.85}
\renewcommand{\bottomfraction}{.7}
\renewcommand{\textfraction}{.15}
\renewcommand{\floatpagefraction}{.66}
\renewcommand{\dbltopfraction}{.66}
\renewcommand{\dblfloatpagefraction}{.66}
\setcounter{topnumber}{9}
\setcounter{bottomnumber}{9}
\setcounter{totalnumber}{20}
\setcounter{dbltopnumber}{9}

@@

\title{Supplementary material for Sequential rank agreement methods for comparison of ranked lists}

\author{CLAUS THORN EKSTRØM$^\ast$, THOMAS ALEXANDER GERDS, ANDREAS KRYGER JENSEN\\[4pt]
%
\textit{%
Biostatistics, Department of Public Health,
University of Copenhagen,
Øster Farimagsgade 5 B, P.O.B. 2099,
DK-1014 Copenhagen K, Denmark}
\\[2pt]
% E-mail address for correspondence
{ekstrom@sund.ku.dk}}

@@

% Running headers of paper:
\markboth%
% First field is the short list of authors
{C. T. Ekstrøm and others}
% Second field is the short title of the paper
{Supplementary material for ``Sequential rank agreement for comparison of ranked lists''}

@@

This document provides proofs of the theorem and corollary shown in the main manuscript.
